Seismic velocities and composition of the Canadian crust

### 2014-04-14 paper/ubcThesisSubmission.tex
thunk

\documentclass[review]{elsarticle}
\usepackage{graphicx,multirow,color,natbib,amssymb}
\usepackage[margin=1in]{geometry}

\makeatletter
\renewcommand\@biblabel[1]{}
\renewenvironment{thebibliography}[1]
     {\section*{\refname}%
      \@mkboth{\MakeUppercase\refname}{\MakeUppercase\refname}%
      \list{}%
           {\leftmargin0pt
            \@openbib@code
            \usecounter{enumiv}}%
      \sloppy
      \clubpenalty4000
      \@clubpenalty \clubpenalty
      \widowpenalty4000%
      \sfcode`\.\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
\makeatother

\journal{Tectonophysics}

\graphicspath{ {figures/} }

\newcommand{\cc}[1]{\textcolor{red}{#1}}

\begin{document}
\begin{frontmatter}

\title{Seismic velocity estimation and the Canadian crust}

\author[ubc]{B. Postlethwaite\corref{cor1}}
\ead{bpostlet@eos.ubc.ca}
\author[ubc]{M. G. Bostock}
\ead{bostock@eos.ubc.ca}
\author[ubc]{N. I. Christensen}
\author[gsc]{D. B. Snyder}

\cortext[cor2]{Corresponding author}

\address[ubc]{Earth and Ocean Sciences, UBC, Vancouver BC, Canada}
\address[gsc]{Geological Survey Canada, Natural Resources Canada, Ottawa ON, Canada}
%% ------------------------------------------------------------------------ %%

\begin{abstract}
It has been suggested that processes driving crustal formation in the Archean and Proterozoic were dissimilar and produced crusts with unique bulk properties and average thicknesses. The calibration of models based on evolving mantle fractionation or mantle convection style require accurate estimates of the geological and geophysical properties of crustal provinces to better constrain the details of crustal formation. Fifteen years of publicly accessible teleseismic data from all available Canadian seismic stations are binned in horizontal slowness and deconvolved into receiver functions. We apply a stacking method to retrieve estimates of the bulk crustal velocity ratio $V_P/V_S$ and thickness $H$ from these data under the assumption of locally 1-D structure. We also investigate modifications to this approach that can allow discrimination of $V_S$ and $V_P$ under certain conditions. Bootstrap error analysis is performed for each station dataset and subsets of these measurements are compared with results for matching stations from previous studies. Analysis of results in conjunction with additional velocity estimates from active source seismic studies and a seismic property database affords improved constraints on bulk geological composition of the Canadian landmass that are used to evaluate competing models of crustal formation.
\end{abstract}

\end{frontmatter}
%% -----------------------------------------------%%
\section{Introduction}

The Canadian continental lithosphere is composed of at least fifteen large geological provinces representing crustal evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the bounding platforms of North America's interior comprise the cratonic core of Laurentia (Hammer et al., 2010). Several distinct geological provinces brought together during a series of Paleoproterozoic orogens make up the Canadian Shield (Hoffman, 1988, Figure 1). The largest of these provinces, and the world's largest Archean craton, is the Superior Province. Bounded by Proterozoic orogenic belts, the western and southern Superior is composed of roughly east-west trending geological subprovinces that include the granite-greenstone terranes of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6 Ga (Thurston et al., 1991). In the northwest of the Superior Province on the eastern shores of the Hudson Bay lies the Nuvvuagittuq greenstone belt that recent geochronological work has dated at ~4.28 Ga making it the only known remnant of Hadean crust preserved on Earth (O'Neil et al., 2011). To the east lies the younger, 1.2-0.98 Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision (Eaton et al., 2006). North of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the western extent of the Churchill Province. Studies of the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga (Thompson et al., 2010). The Slave Province is a relatively small Archean granite-greenstone block lying to the northwest of the Churchill province. Its western portion is the 4.0 Ga remnant of a larger Archean continent, whereas the eastern Slave forms a younger (2.6 Ga) paired accretionary prism and island arc (Kusky, 1989). To the west of the Superior is the extensive Interior Platform overlying the southern extensions of the Trans-Hudson Orogen, Hearne Domain and Alberta Orogen. The Interior Platform is largely Archean reworked during the Paleoproterozoic with bore-hole samples indicating amphibolite to granulite-grade Proterozoic metamorphism (Hammer et al., 2010). Extending west from the Interior platform, the Western Cordillera developed in a passive margin setting on and adjacent to late Archean to Paleoproterozoic crystalline basement. Between the middle of the Jurassic and Paleocene intense metamorphic, plutonic and uplifting events occurred as the result of Paleozoic and younger intraoceanic-arc accretionary events. The Coast Mountains evolved during the late Cretaceous as a wide magmatic arc composed largely of granitic rock (Clowes et al, 1995). Although there is considerable variation in structure and tectonic history, roughly half of the Cordillera is underlain by Proterozoic-margin crust that forms a crustal-scale d\'{e}collement beneath interleaved and duplexed crust (Hammer et al., 2010). The rich tectonic tapestry of the Canadian continental crust affords an excellent opportunity to study variations in bulk crustal properties with geological age.

Our knowledge of the continental crust in Canada, in particular its thickness and structure in depth, has advanced greatly through Lithoprobe, a now complete national geoscience project spearheaded by active-source seismic profiling (Clowes et al., 1999). The full suite of Lithoprobe seismic profiles form a near-continuous transect from the Pacific to the Atlantic Ocean, and has enabled many insights into crustal structure and genesis. Recent summary studies utilizing Lithoprobe data include the analysis of {\it Eaton} (2005) which considers seismic observations in light of several hypotheses for the origin of the continental Moho: relic Moho, underplating, metamorphic front and regional d\'{e}collement. Seismic evidence of preserved island arc Moho in Cascadia supports the first hypothesis whereas the presence of high-velocity, lower-crustal xenoliths in southern Alberta favours thick mafic underplating. Eclogite mineral assemblages forming a deep crustal root below the Grenville Front lends support to the third hypothesis, that at least in some places, continental Moho forms by large-scale eclogitization of the lower crust. Finally, seismic imaging of subhorizontal reflections separated by dipping reflection fabrics beneath the southern Canadian Cordillera suggests decoupling of the upper crust from the mantle during deformation and is taken as evidence for the fourth hypothesis, that an overprinted Moho may serve as a subhorizontal regional d\'ecollement. A subsequent study by {\it Cook et al.} (2009) argues that the continental Moho has varied and complex origins and that its imprint evolves through the effects of temperature, pressure, deformation and fluid migration.

In another recent summary study of Lithoprobe data, Hammer and Clowes (2011) discuss broad structural characteristics of the Canadian crust and note that Archean cratons and their boundaries contain structures similar to those found in Proterozoic and Phanerozoic orogens. They also discuss the lack of evidence for significant crustal disruption from magmatic intrusion related to subduction, post-orogenic extension or orogenic collapse. The authors remark upon the striking uniformity in Moho depth that is imaged between 33 km and 43 km depth along 20,000 km of Lithoprobe transects, despite the diversity of overlying crust. Both {\it Cook et al.} (2009) and {\it Hammer and Clowes} (2011) raise the possibility that, in some circumstances, the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary but rather eclogitized lower crust that is seismically indistinguishable from the upper mantle. If this is the case, the seismic Moho would represent the top of a metamorphic front and would be shallower than the crust-mantle boundary.

The Lith5.0 revised crustal model presented by {\it Perry et al.} (2002) was developed from crustal thickness estimates taken from Lithoprobe seismic profiles. This work updates the global Crust5.1 model (Mooney et al, 1998) and introduces broad crustal features of the Canadian crust such as relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick ($>$42.5 km) Phanerozoic crust in the south and west of the continental craton.

In this study, we examine the additional constraints that analyses of teleseismic receiver functions place upon Canadian bulk crustal properties when combined with available active source results. It is widely appreciated that receiver functions are sensitive primarily to crustal thickness and Poisson's ratio (or equivalently $V_P/V_S$) (Zhu and Kanomori, 2000), and that the latter quantity provides some ability to discriminate among possible bulk compositions (Zandt and Ammon, 1995). The constraint on Poisson's ratio and the greater averaging performed by low frequency (0.04-3 Hz) scattered teleseismic waves versus high frequency (10-40 Hz) {\it P}-waves employed in active source studies render the two techniques somewhat complementary. Our data set comprises the majority of teleseismic data available for the Canadian landmass including both permanent and temporary deployments. Many of the temporary deployments are described by {\it Bostock et al.} (2010). Three previous studies employing similar methodologies but with more regional emphases deserve special mention: an examination of crustal thickness and $V_P/V_S$ variation in the Grenville Orogen (Eaton et al., 2005), a structural and compositional characterization of the Superior Province (Darbyshire et al. 2006) and a crustal evolution analysis of the Hudson Bay region (Thompson et al., 2010). Results from these studies will serve as benchmarks for our expanded effort. In the following sections we describe the data and methods employed to interrogate the Canadian crust. We proceed to present the major results beginning with a comparison between our seismic parameter estimates and the estimates from three studies mentioned above for identical seismic stations. We then present the results of a modified approach to our parameter estimation method and follow this with a presentation of the resulting crustal property estimates.  First presented are the average crustal discontinuity and structural patterns across Canada. Second, we take a tour through those geological provinces for which sufficient data exists, focusing on regional variation and characterization. We conclude with a discussion on the implications for our understanding of the composition and evolution of the Canadian crust.

%% -----------------------------------------------%%

\section{Data and methods}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{stationMap.pdf}
  \caption{Map showing major Canadian geological provinces and seismic stations contributing to the data discussed in this study.}
  \label{map:stationMap}
\end{figure}

@@

\subsection{Teleseismic Data Set}
The primary data utilized in this study are computed from teleseismic {\it P} wave seismograms originating from more than 700 earthquake sources. These earthquakes occurred between 2000 and 2012 and were recorded on subsets of 343 broadband seismic stations distributed across Canada. Seismic stations are selected from all available regional and national networks including CNSN, POLARIS, FedNor and Chasme. Seismograms are included for analysis if they fall within a 30$^\circ$-100$^\circ$ epicentral distance window. Quality control is performed by admitting only those seismograms with sufficiently high signal to noise ratios to permit visual observation of the dominant {\it P} wave energy and with sufficiently impulsive first arrivals to allow arrival onsets to be accurately measured. After selection and filtering more than 80,000 events are available for further processing. Seismic stations contributing to the data utilized in this study are shown in figure \ref{map:stationMap}.

The first stage of processing requires the transformation of teleseismic data into receiver functions. In general terms, this transformation involves deconvolving an approximation of the earthquake source from horizontal recordings of ground motion (Langston, 1979). The resulting waveforms contain discrete pulses corresponding to {\it S} waves scattered from subsurface discontinuities including the base of the crust, or Moho. Confident identification of these peaks is required to estimate crustal parameters, and improved results can be obtained by transforming the displacement seismogram into their {\it P} and {\it S} contributions based on a 1-D model of propagation. This procedure is accomplished by first rotating the N and E coordinates into radial and transverse directions and then performing a wave field decomposition using the radial and vertical channels (Kennet, 1991). The direct {\it P} arrival of the signal on the resulting {\it P} wave component is used as an approximation to the source function as {\it P} component impulse response for teleseismic P approximates a delta function. This windowed source estimate is deconvolved from the {\it S} wave component computed from the wave field decomposition.

We employ an $L_2$, multichannel, frequency domain approach to perform the deconvolution that has the advantage of computational efficiency and does not require {\it a priori} assumptions concerning the noise levels in the data. More specifically, we employ a simultaneous deconvolution of $N$ seismograms sharing a similar slowness to compute a single impulse response or receiver function $r(t)$.

@@

\noindent where $F^{-1}$ is the inverse Fourier transform, $S_n$ represents the $n^{th}$ {\it S} wave seismogram, $P_n$ is the corresponding windowed {\it P} wave seismogram, $^*$ denotes complex conjugation and $\delta$ is the regularization parameter controlling the trade off between model smoothness and data misfit. The parameter $\delta$ which is chosen programatically by minimizing the general cross validation function $GCV(\delta)$ is given by

@@

\noindent and $\omega$ is the frequency bin in the discrete Fourier transform. All resulting receiver functions, $r(t)$, are filtered between 0.04Hz and 3.0Hz.

\subsection{Vp/Vs method} \label{section:VpVsMethod}

\begin{figure}
  \centering
    \includegraphics[width=0.7\textwidth]{reflectedPhases.pdf}
  \caption{Schematic diagram illustrating geometry of phases for the velocity contrast representing the Moho}
  \label{fig:reflectedPhases}
\end{figure}

>>>>

%%
%% This is file `ubcsample.tex',
%% generated with the docstrip utility.
%
% The original source files were:
%
% ubcthesis.dtx  (with options: `ubcsampletex')
%%
%% This file was generated from the ubcthesis package.
%% --------------------------------------------------------------
%% NEED TO FIX: Envelope, ensure abs value --- add to figure caption
%% Lith figure, tighten range.
%% Rework final geochron conclusion. Data is weak for current claims
%% figure -> Figure
\documentclass[msc,oneside]{ubcthesis}

\usepackage{graphicx,multirow,color,amssymb}
\usepackage[margin=1in]{geometry}
\graphicspath{ {figures/} }
\newcommand{\cc}[1]{\textcolor{red}{#1}}

%******** afterpage ***************************
% This package allows you to issue commands at the end of the current
% page.  A good use for this is to use the command
% \afterpage{\clearpage} right after a figure.  This will cause the
% figure to be inserted on the page following the current one (or on
% the current page if it will fit) but will not break the page in the
% middle.
\usepackage{afterpage}

%******** float *********************************
% This package allows you to customize the style of
% "floats"---floating objects such as figures and tables.  In
% addition, it allows you to define additional floating objects which
% may be included in a list similar to that produces by \listoftables
% and \listoffigures.  Common uses include introducing floats for
% programs and other code bits in Compute Science and Chemical Schema.
\usepackage{float}

%******** tocloft *******************************
% This package allows you to customize and define custom lists such
% as a list of programs or Chemical Scheme.  Note: if you use the
% subfigure package, you must specify that you do as an option here.
% The title option uses the default formatting.  We do not use this
% here as the default formatting is acceptable.  Use the float
% package instead unless you need the extra formatting control
% provided by tocloft.
%\usepackage[subfigure, titles]{tocloft}

%******** alltt *********************************
% The alltt package allows you to include files and have them
% formatted in a verbatim fashion.  This is useful for including
% source code from an additional file.
%\usepackage{alltt}

%******** listings ******************************
% The listings package may be used to include chunks of source code
% and has facilities for pretty-printing many languages.
%\usepackage{listings}

%******** longtable *****************************
% The longtable package allows you to define tables that span
% multiple pages.
% \usepackage{longtable}

%******** graphics and graphicx *****************
% This allows you to include encapsulated postscript files.  If you
% don't have this, comment the \includegraphics{} line following the
% comment "%includegraphics" later in this file.
% \usepackage{graphicx}

%******** subfigure *****************************
% The subfigure package allows you to include multiple figures and
% captions within a single figure environment.
%\usepackage{subfigure}

%******** here **********************************
% The here package gives you more control over the placement of
% figures and tables.  In particular, you can specify the placement
% "H" which means "Put the figure here" rather than [h] which means
% "I would suggest that you put the figure here if you think it looks
% good."
%\usepackage{here}

%******** natbib ********************************
% This is a very nice package for bibliographies.  It includes options
% for sorting and compressing bibliographic entries.
\usepackage[numbers,sort&compress]{natbib}

%******** psfrag ******************************
% This allows you to replace text in postscript pictures with formated
% latex text.  This allows you to use math in graph labels
% etc. Uncomment the psfrag lines following the "%psfrag" comment
% later in this file if you don't have this package.  The replacements
% will only be visible in the final postscript file: they will be
% listed in the .dvi file but not performed.
\usepackage{psfrag}

%******** hyperref *****************************
% Please read the manual:
% http://www.tug.org/applications/hyperref/manual.html
%
% This adds hyperlinks to your document: with the right viewers (later
% versions of xdvi, acrobat with pdftex, latex2html etc.) this will
% make your equation, figure, citation references etc. hyperlinks so
% that you can click on them.  Also, your table of contents will be
% able to take you to the appropriate sections.  In the viewers that
% support this, the links often appear with an underscore.  This
% underscore will not appear in printed versions.
%
% Note: if you do not use the hypertex option, then the dvips driver
% may be loaded by default.  This will cause the entries in the list
% of figures and list of tables to be on a single line because dvips
% does not deal with hyperlinks on broken lines properly.
%
% NOTE: HYPERREF is sensitive to the ORDER in which it is LOADED.
% For example, it must be loaded AFTER natbib but BEFORE newly
% defined float environments.  See the README file with the hyperref
% for some help with this.  If you have some very obscure errors, try
% first disabling hyperref.  If that fixes the problem, try various
% orderings.
%
% Note also that there is a bug with versions before 2003/11/30
% v6.74m that cause the float package to not function correctly.
% Please ensure you have a current version of this package.  A
% warning will be issued if you leave the date below but do not have
% a current version installed.
%
% Some notes on options: depending on how you build your files, you
% may need to choose the appropriate option (such as [pdftex]) for the
% backend driver (see the hyperref manual for a complete list).  Also,
% the default here is to make links from the page numbers in the table
% of contents and lists of figures etc.  There are other options:
% excluding the [linktocpage] option will make the entire text a
% hyperref, but for some backends will prevent the text from wrapping
% which can look terrible.  There is a [breaklinks=true] option that
% will be set if the backend supports (dvipdfm for example supports
% it but does not work with psfrag.)
%
% Finally, there are many options for choosing the colours of the
% links.  These will be included by default in future versions but
% you should probably consider changing some now for the electronic
% version of your thesis.
\usepackage[unicode=true,
  linktocpage,
  linkbordercolor={0.5 0.5 1},
  citebordercolor={0.5 1 0.5},
  linkcolor=blue]{hyperref}

% If you would like to compile this sample thesis without the
% hyperref package, then you will need to comment out the previous
% \usepackage command and uncomment the following command which will
% put the URL's in a typewriter font but not link them.
%\newcommand\url[1]{\texttt{#1}}

%******** setspace *******************************
% The setspace package allows you to manually set the spacing of the
% file.  UBC may require 1.5 spacing for microfilming of theses.  In
% this case you may obtain this by including this package and issuing
% one of the following commands:
%\usepackage{setspace}
%\singlespacing
%\onehalfspacing
%\doublespacing

% These commands are optional.  The defaults are shown.  You only
% need to include them if you need a different value
\institution{The University Of British Columbia}

% If you are at the Okanagan campus, then you should specify these
% instead.
%\faculty{The College of Graduate Studies}
%\institutionaddress{Okanagan}
\faculty{The Faculty of Graduate Studies}
\institutionaddress{Vancouver}

% You can issue as many of these as you have...
\previousdegree{B.Sc., The University of British Columbia, 1999}
\previousdegree{M.Sc., The University of British Columbia, 2001}
\previousdegree{Ph.D., Massachusetts Institute of Technology, 2005}

% You can override the option setting here.
% \degreetitle{Jack of All Trades}

% These commands are required.
\title{Seismic velocities and composition of the Canadian crust}
\author{Ben Postlethwaite}
\copyrightyear{2014}
\submitdate{\monthname\ \number\year} % The "\ " is required after
                                      % \monthname to prevent the
                                      % command from eating the space.
\program{GeoPhysics}

% These commands are presently not required for UBC theses as the
% advisor's name and title are not presently required anywhere.
%\advisor{Ariel R.~Zhitnitsky}
%\advisortitle{Professor of Physics}

% One might want to override the format of the section and chapter
% numbers.  This shows you how to do it.  Note that the current
% format is acceptable for submission to the FoGS: If you wish to modify
% these, you should check with the FoGS explicity. prior to making
% the modifications.
\renewcommand\thepart         {\Roman{part}}
\renewcommand\thechapter      {\arabic{chapter}}
\renewcommand\thesection      {\thechapter.\arabic{section}}
\renewcommand\thesubsection   {\thesection.\arabic{subsection}}
\renewcommand\thesubsubsection{\thesubsection.\arabic{subsubsection}}
\renewcommand\theparagraph    {\thesubsubsection.\arabic{paragraph}}
\renewcommand\thesubparagraph {\theparagraph.\arabic{subparagraph}}

\setcounter{tocdepth}{2}
\setcounter{secnumdepth}{2}

% Here is an example of a "Program" environment defined with the
% "float" package.  The list of programs will be stored in the file
% ubcsample.lop and the numbering will start with the chapter
% number.  The style will be "ruled".
\floatstyle{ruled}
\newfloat{Program}{htbp}{lop}[chapter]

% Here is the start of the document.
\begin{document}

%% This starts numbering in Roman numerals as required for the thesis
%% style and is mandatory.
\frontmatter

%%% The order of the following components should be preserved.  The order
%%% listed here is the order currently required by FoGS:        \\
%%% Title (Mandatory)                                           \\
%%% Preface (Manditory if any collaborator contributions)       \\
%%% Abstract (Mandatory)                                        \\
%%% List of Contents, Tables, Figures, etc. (As appropriate)    \\
%%% Acknowledgements (Optional)                                 \\
%%% Dedication (Optional)                                       \\

\maketitle                      %% Mandatory
\begin{abstract}                %% Mandatory -  maximum 350 words
  It has been suggested that processes driving crustal formation in the Archean and Proterozoic were dissimilar and produced crusts with unique bulk properties and average thicknesses. The calibration of models based on evolving mantle fractionation or mantle convection style require accurate estimates of the geological and geophysical properties of crustal provinces to better constrain the details of crustal formation. Fifteen years of publicly accessible teleseismic data from all available Canadian seismic stations are binned in horizontal slowness and deconvolved into receiver functions. We apply a stacking method to retrieve estimates of the bulk crustal velocity ratio $V_P/V_S$ and thickness $H$ from these data under the assumption of locally 1-D structure. We also investigate modifications to this approach that can allow discrimination of $V_S$ and $V_P$ under certain conditions. Bootstrap error analysis is performed for each station dataset and subsets of these measurements are compared with results for matching stations from previous studies. Analysis of results in conjunction with additional velocity estimates from active source seismic studies and a seismic property database affords improved constraints on bulk geological composition of the Canadian landmass that are used to evaluate competing models of crustal formation.

  Ben Postlethwaite
  post.ben.here@gmail.com
\end{abstract}

\chapter{Preface} % Manditory if any of the conditions are met

You must include a preface if any part of your research was partly or
wholly published in articles, was part of a collaboration, or required
the approval of UBC Research Ethics Boards.

The Preface must include the following:

\begin{itemize}
\item A statement indicating the relative contributions of all
  collaborators and co-authors of publications (if any), emphasizing
  details of your contribution, and stating the proportion of research
  and writing conducted by you.
\item A list of any publications arising from work presented in the
  dissertation, and the chapter(s) in which the work is located.
\item The name of the particular UBC Research Ethics Board, and the
  Certificate Number(s) of the Ethics Certificate(s) obtained, if
  ethics approval was required for the research.
\end{itemize}

%%% Sections and subsections etc. in the Preface should in general
%%% not be listed in the table of contents, so use the starred form
%%% of \section etc.
\section*{Examples}
Chapter~\ref{cha:apple_ref} is based on work conducted in UBC's Maple
Syrup Laboratory by Dr. A.  Apple, Professor B. Boat, and Michael
McNeil Forbes. I was responsible for tapping the trees in forests X
and Z, conducted and supervised all boiling operations, and performed
frequent quality control tests on the product.

A version of chapter~\ref{cha:apple_ref} has been
published~\cite{Apple:2010}. I conducted all the testing and wrote
most of the manuscript. The section on ``Testing Implements'' was
originally drafted by Boat, B.  Check the first pages of this
chapter to see footnotes with similar information.

Note that this preface must come before the table of contents.  Note
also that this section ``Examples'' should not be listed in the table
of contents, so we have used the starred form: \verb|\section*{Example}|.

\tableofcontents                %% Mandatory
\listoftables                   %% Mandatory if thesis has tables
\listoffigures                  %% Mandatory if thesis has figures

\chapter{Acknowledgements}      %% Optional
This is the place to thank professional colleagues and people who have
given you the most help during the course of your graduate work.

% Now regular page numbering begins.
\mainmatter

% Parts are the largest structural units, but are optional.
%\part{Thesis}

% Chapters are the next main unit.
\chapter{Introduction}

% Sections are a sub-unit
\section{Geology}
The Canadian continental lithosphere is composed of at least fifteen large geological provinces representing crustal evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the bounding platforms of North America's interior comprise the cratonic core of Laurentia (Hammer et al., 2010). Several distinct geological provinces brought together during a series of Paleoproterozoic orogens make up the Canadian Shield (Hoffman, 1988, Figure 1). The largest of these provinces, and the world's largest Archean craton, is the Superior Province. Bounded by Proterozoic orogenic belts, the western and southern Superior is composed of roughly east-west trending geological subprovinces that include the granite-greenstone terranes of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6 Ga (Thurston et al., 1991). In the northwest of the Superior Province on the eastern shores of the Hudson Bay lies the Nuvvuagittuq greenstone belt that recent geochronological work has dated at ~4.28 Ga making it the only known remnant of Hadean crust preserved on Earth (O'Neil et al., 2011). To the east lies the younger, 1.2-0.98 Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision (Eaton et al., 2006). North of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the western extent of the Churchill Province. Studies of the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga (Thompson et al., 2010). The Slave Province is a relatively small Archean granite-greenstone block lying to the northwest of the Churchill province. Its western portion is the 4.0 Ga remnant of a larger Archean continent, whereas the eastern Slave forms a younger (2.6 Ga) paired accretionary prism and island arc (Kusky, 1989). To the west of the Superior is the extensive Interior Platform overlying the southern extensions of the Trans-Hudson Orogen, Hearne Domain and Alberta Orogen. The Interior Platform is largely Archean reworked during the Paleoproterozoic with bore-hole samples indicating amphibolite to granulite-grade Proterozoic metamorphism (Hammer et al., 2010). Extending west from the Interior platform, the Western Cordillera developed in a passive margin setting on and adjacent to late Archean and Paleoproterozoic crystalline basement. Between the middle of the Jurassic and Paleocene intense metamorphic, plutonic and uplifting events occurred as the result of Paleozoic and younger intraoceanic-arc accretionary events. The Coast Mountains evolved during the late Cretaceous as a wide magmatic arc composed largely of granitic rock (Clowes et al, 1995). Although there is considerable variation in structure and tectonic history, roughly half of the Cordillera is underlain by Proterozoic-margin crust that forms a crustal-scale d\'{e}collement beneath interleaved and duplexed crust (Hammer et al., 2010). The rich tectonic tapestry of the Canadian continental crust affords an excellent opportunity to study variations in bulk crustal properties with geological age.

\section{Previous Work}
Our knowledge of the continental crust in Canada, in particular its thickness and structure in depth, has advanced greatly through {\sc Lithoprobe}, a now complete national geoscience project spearheaded by active-source seismic profiling (Clowes et al., 1999). The full suite of {\sc Lithoprobe} seismic profiles form a near-continuous transect from the Pacific to the Atlantic Ocean, and has enabled many insights into crustal structure and genesis. Recent summary studies utilizing {\sc Lithoprobe} data include the analysis of Eaton (2005) which considers seismic observations in light of several hypotheses for the origin of the continental Moho: relic Moho, underplating, metamorphic front and regional d\'{e}collement. Seismic evidence of preserved island arc Moho in Cascadia supports the first hypothesis whereas the presence of high-velocity, lower-crustal xenoliths in southern Alberta favours thick mafic underplating. Eclogite mineral assemblages forming a deep crustal root below the Grenville Front lend support to the third hypothesis, that at least in some places, continental Moho forms by large-scale eclogitization of the lower crust. Finally, seismic imaging of subhorizontal reflections separated by dipping reflection fabrics beneath the southern Canadian Cordillera suggests decoupling of the upper crust from the mantle during deformation and is taken as evidence for the fourth hypothesis, that an overprinted Moho may serve as a subhorizontal regional d\'ecollement. A subsequent study by {\it Cook et al.} (2009) argues that the continental Moho has varied and complex origins and that its imprint evolves through the effects of temperature, pressure, deformation and fluid migration.

In another recent summary study of {\sc Lithoprobe} data, Hammer and Clowes (2011) discuss broad structural characteristics of the Canadian crust and note that Archean cratons and their boundaries contain structures similar to those found in Proterozoic and Phanerozoic orogens. They also discuss the lack of evidence for significant crustal disruption from magmatic intrusion related to subduction, post-orogenic extension or orogenic collapse. The authors remark upon the striking uniformity in Moho depth that is imaged between 33 km and 43 km depth along 20,000 km of {\sc Lithoprobe} transects, despite the diversity of overlying crust. Both {\it Cook et al.} (2009) and {\it Hammer and Clowes} (2011) raise the possibility that, in some circumstances, the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary but rather eclogitized lower crust that is seismically indistinguishable from the upper mantle. If this is the case, the seismic Moho would represent the top of a metamorphic front and would be shallower than the crust-mantle boundary.

The Lith5.0 revised crustal model presented by {\it Perry et al.} (2002) was developed from crustal thickness estimates taken from {\sc Lithoprobe} seismic profiles. This work updates the global Crust5.1 model (Mooney et al, 1998) and emphasizes broad crustal features of the Canadian crust such as relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick ($>$42.5 km) Phanerozoic crust in the south and west of the continental craton.

\section{Summary}
In this study, we examine the additional constraints that analyses of teleseismic receiver functions place upon Canadian bulk crustal properties when combined with available active source results. It is widely appreciated that receiver functions are sensitive primarily to crustal thickness and Poisson's ratio (or equivalently $V_P/V_S$) (Zhu and Kanomori, 2000), and that the latter quantity provides some ability to discriminate among possible bulk compositions (Zandt and Ammon, 1995). The constraint on Poisson's ratio and the greater averaging performed by low frequency (0.04-3 Hz) scattered teleseismic waves versus high frequency (10-40 Hz) {\it P}-waves employed in active source studies render the two techniques somewhat complementary. Our data set comprises the majority of teleseismic data available for the Canadian landmass including both permanent and temporary deployments. Many of the temporary deployments are described by Bostock et al. (2010). Three previous studies employing similar methodologies but with more regional emphases deserve special mention: an examination of crustal thickness and $V_P/V_S$ variation in the Grenville Orogen (Eaton et al., 2005), a structural and compositional characterization of the Superior Province (Darbyshire et al. 2006) and a crustal evolution analysis of the Hudson Bay region (Thompson et al., 2010). Results from these studies will serve as benchmarks for our expanded effort.

In the following sections we describe the data and methods employed to interrogate the Canadian crust. We proceed to present the major results beginning with a comparison between our seismic parameter estimates and the estimates from three studies mentioned above for identical seismic stations. We then present the results of a modified approach to our parameter estimation method and follow this with a presentation of the resulting crustal property estimates.  First presented are the average crustal discontinuity and structural patterns across Canada. Second, we take a tour through those geological provinces for which sufficient data exists, focusing on regional variation and characterization. We conclude with a discussion on the implications for our understanding of the composition and evolution of the Canadian crust.

\chapter{Data and Methods}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{stationMap.pdf}
  \caption[Seismic broadband station locations]{Map showing major Canadian geological provinces and seismic stations contributing to the data discussed in this study.}
  \label{map:stationMap}
\end{figure}

@@

\section{Teleseismic Data Set}
The primary data utilized in this study are RF processed teleseismic {\it P} wave seismograms originating from more than 700 earthquake sources. These earthquakes occurred between 2000 and 2012 and were recorded on subsets of 343 broadband seismic stations distributed across Canada. Seismic stations are selected from all available regional and national networks including CNSN, POLARIS, FedNor and Chasme. Seismograms are included for analysis if they fall within a 30$^\circ$-100$^\circ$ epicentral distance window. Quality control is performed by admitting only those seismograms with sufficiently high signal to noise ratios to permit visual observation of the dominant {\it P} wave energy and with sufficiently impulsive first arrivals to allow arrival onsets to be accurately measured. After selection and filtering more than 80,000 seismograms are available for further processing. Seismic stations contributing to the data utilized in this study are shown in figure \ref{map:stationMap}.

The first stage of processing requires the transformation of teleseismic data into receiver functions. In general terms, this transformation involves deconvolving an approximation of the earthquake source from horizontal recordings of ground motion (Langston, 1979). The resulting waveforms contain discrete pulses corresponding to {\it S} waves scattered from subsurface discontinuities including the base of the crust, or Moho. Confident identification of these peaks is required to estimate crustal parameters, and improved results can be obtained by transforming the displacement seismogram into its {\it P} and {\it S} contribution based on a 1-D model of propagation. This procedure is accomplished by first rotating the N and E coordinates into radial and transverse directions and then performing a wave field decomposition using the radial and vertical channels (Kennett, 1991). The direct {\it P} arrival of the signal on the resulting {\it P} wave component is used as an approximation to the source function because the {\it P} component impulse response for teleseismic P approximates a delta function. This windowed source estimate is deconvolved from the {\it S} wave component computed from the wave field decomposition.

We employ an $L_2$, multichannel, frequency domain approach to perform the deconvolution that has the advantage of computational efficiency and does not require {\it a priori} assumptions concerning noise levels in the data. More specifically, we employ a simultaneous deconvolution of $N$ seismograms sharing a similar slowness to compute a single impulse response or receiver function $r(t)$.

@@

\noindent where $F^{-1}$ is the inverse Fourier transform, $S_n$ represents the $n^{th}$ {\it S} wave seismogram, $P_n$ is the corresponding windowed {\it P} wave seismogram, $^*$ denotes complex conjugation and $\delta$ is the regularization parameter controlling the trade off between model smoothness and data misfit. The parameter $\delta$ which is chosen programatically by minimizing the general cross validation function $GCV(\delta)$ (Golub et al., 1979) is given by

@@

\noindent and $\omega_m$ is the $m^{th}$ discrete frequency. All resulting receiver functions, $r(t)$, are filtered between 0.04Hz and 3.0Hz.

\section{Vp/Vs method} \label{section:VpVsMethod}

\begin{figure}
  \centering
    \includegraphics[width=0.7\textwidth]{reflectedPhases.pdf}
  \caption[Teleseismic reflection phase diagram]{Schematic diagram illustrating geometry of phases for the velocity contrast representing the Moho}
  \label{fig:reflectedPhases}
\end{figure}
